pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-20 article.tex
Process feedback @Giappo

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium (as known as "early burst") [?Harmon \giovanni{is it this one? https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2010.01025.x}],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny.

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
\giovanni{Here it could be useful to add a figure that explains the pipeline, a bit like the one I have on my desk}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;babette;\giovanni{babette or mcbette?}, that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

We start with an idealised \giovanni{I still don't get why we need an "idealized" tree.} birth-death tree of five taxa and a crown age of 10:

% Need idealised birth-death tree
% From https://github.com/richelbilderbeek/pirouette_article/issues/1
% and https://github.com/richelbilderbeek/pirouette/issues/35
\begin{lstlisting}[language=R, floatplacement=H, frame=single]
phylogeny <- ape::read.tree(text = "((A:7.1,(B:2.2,C:2.2):4.9):2.9,(D:4.6,E:4.6):5.4):0;")
\end{lstlisting}

@@

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model, clock model and tree prior. \giovanni{Here you present the "alignment generative model" and the "inference model". These are general concepts, as they are used in all the experiments. I propose to define them from the start, for example in the "description" section; we can even refer to them in the figure I propose to add, so to make it clearer to the reader. It can be done in this way:
A "model" is always defined as the collection of three elements: a site model, a clock model and a tree prior. We have two kind of models: generative and inference. The first is used to simulate alignments, the second is used to produce a posterior. Generative models always use a (constant-rate) birth-death tree prior.}
If we know how the generative model of the alignment, as we did above,
we can re-use that information. We do still need to specify 
a generative tree prior, which is the tree prior underlying the phylogeny.
In this case, we'll pick a (constant-rate) birth-death model:

@@

\richel{Update figure 2 to use idealized tree}
\giovanni{Check "pirouettepackage" repo. I uploaded a function to create the idealized tree that you might like.}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny using the same 
    diversification model as it was generated by
    \giovanni{If we define generative and inference model from the start we can rephrase this caption in a better way, referring to terms that the reader is already familiar to}
  }
\end{figure}

@@

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence to be used
in the Bayesian inference. By default, \verb;pirouette; will try
all combinations of site models, clock models and tree priors.\giovanni{again, if we give a name to this triplet, it could be good to call it by its name. We can still recall its composition in brackets, if we think it could be useful}

@@

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
an idealized \giovanni{I don't think this is true. A twin tree is generated randomly. If we use the method "maxlikelihood" (see the function "createbdtree") the tree will be chosen in such a way it will maximizing the likelihood given the parameters we estimated from the original tree, but I would not call it "idealized". Furthermore I think it's not obvious to use the "maxlikelihood" method, as in general the "ideal" scenario should be established using a distribution of trees. A single tree is always the result of a stochastic process.} birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise?
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. \giovanni{again, I don't believe that using an "idealized" tree is the best thing to do. I would rather use more "true trees".}
  }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\giovanni{We need to discuss the results we got from the 3 experiments.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium (as known as "early burst") [\cite{harmon2010early}],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given reconstructed phylogeny. For this measurement,
two models need to be specified. Both models are defined as a combination of 
site model, clock model and a tree prior. The two kind of models are a 
generative and inference model. The generative model is used to simulate 
alignments, the inference model is used to produce a posterior. 
Generative models always use a (constant-rate) birth-death tree prior.

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
\richel{figure to pipeline here}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; \cite{mcbette}, that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

We start with an idealised 
birth-death tree of five taxa and a crown age of 10.
By idealized, we mean a phylogeny that is easiest to recover
by the Bayesian inference. A phylogeny that is easy to recover has
the most common branch length distribution imaginable. We
use an idealised tree, as we want to measure the baseline noise; that is,
the lowest error that BEAST2 can possibly make.

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
phylogeny <- ape::read.tree(text = "((A:7.1,(B:2.2,C:2.2):4.9):2.9,(D:4.6,E:4.6):5.4):0;")
\end{lstlisting}

@@

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model, clock model and tree prior. 
If we know how the generative model of the alignment, as we did above,
we can re-use that information. We do still need to specify 
a generative tree prior, which is the tree prior underlying the phylogeny.
In this case, we'll pick a (constant-rate) birth-death model:

@@

\richel{Update figure 2 to use idealized tree}
\giovanni{
  Check "pirouettepackage" repo. 
  I uploaded a function to create the idealized tree that you might like.
}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    when the generative and inference model are the same
  }
\end{figure}

@@

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence to be used
in the Bayesian inference. By default, \verb;pirouette; will use
its full arsenal of 40 inference models, which are all combinations of 4 site 
models, 2 clock models and 5 tree priors.

@@

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
an idealized 
\giovanni{
  I don't think this is true. A twin tree is generated randomly. 
  If we use the method "maxlikelihood" (see the function "createbdtree") 
  the tree will be chosen in such a way it will maximizing the likelihood 
  given the parameters we estimated from the original tree, 
  but I would not call it "idealized". Furthermore I think it's not obvious to 
  use the "maxlikelihood" method, as in general the "ideal" scenario should be 
  established using a distribution of trees. A single tree is always the result 
  of a stochastic process.
}
\richel{we'll discuss this today} 
birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise?
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. 
    \giovanni{again, I don't believe that using an "idealized" tree is the best 
      thing to do. I would rather use more "true trees".}
    \richel{we'll discuss this today}
  }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\giovanni{We need to discuss the results we got from the 3 experiments.}
\richel{Only iff the journal requires this}
